---
title: "Pan-cancer analysis"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
theme_set(ggpubr::theme_pubclean())

inputDir <- '/data/local/buyar/arcas/subtyping_paper/pancancer/output/'
assay_dir <- '/data/local/buyar/arcas/subtyping_paper/pancancer/assays/'
surv <- do.call(rbind, readRDS('/data/local/buyar/pancancer_diagnostics/data/GDCdata_processed2/TCGA.surv.RDS'))
projectDict <- sapply(yaml::read_yaml('/data/local/buyar/arcas/subtyping_paper/data/projects.dict.yaml'), 
                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
tcga_subtypes <- process_tcga_subtypes()
```

```{r import_lfs}
LFs <- sapply(simplify = F, c('pca_factors', 'maui_factors', 'mofa_factors'), function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'factors.csv$', full.names = T)
  LFs <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- dt$V1
    return(M)
  })
  names(LFs) <- unlist(lapply(strsplit(basename(names(LFs)), split = '\\.'), function(x) paste(x[1:2], collapse = '.')))
  return(LFs)
})
```

```{r import_feature_weights}
FWs <- sapply(simplify = F, c('pca_factors', 'maui_factors', 'mofa_factors'), function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'feature_weights.csv$', full.names = T)
  W <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- gsub("^assay: ", "", dt$V1) # required only for maui
    return(M)
  })
  names(W) <- unlist(lapply(strsplit(basename(names(W)), split = '\\.'), 
                            function(x) paste(x[1:2], collapse = '.')))
  return(W)
})
```


```{r}
# Define a function that takes as input LFs and some patient labels (clinical/cancer type/subtype)
# and evaluates predictability of the labels using the LFs
predict_labels <- function(LFs, surv, tcga_subtypes, 
                           project, datatype, label.type, ...) {
  source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
  predict_stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
    require(data.table)
    message("predicting ",label.type," using ",tool)
    experiment <- paste(project, datatype, sep = '.')
    M <- LFs[[tool]][[experiment]]
    if(label.type == 'cancer_type') {
      labels <- data.frame('cancer_type' = surv[match(rownames(M), bcr_patient_barcode)]$project, 
                           row.names = rownames(M))
    } else if (label.type == 'subtype') {
      labels <- data.frame('subtype' = tcga_subtypes[match(rownames(M),
                                                           bcr_patient_barcode)]$Subtype_Selected, 
                           row.names = rownames(M))
    } else if (label.type == 'clinical') {
      labels <- process_clinical_covariates(rownames(M))
    }
    dt <- do.call(rbind, lapply(colnames(labels), function(x) {
      l <- labels[,x]
      if(is.numeric(l)){
        # categorize samples by intervals
        l <- as.character(cut_interval(l, 10))
      }
      l[is.na(l)] <- 'Undefined'
      message("Predicting ",x, " for experiment ",experiment)
      dt <- suppressMessages(predict_cluster_labels(M, l, ...))
      dt$label.type <- x
      dt$project <- project
      dt$datatypes <- datatype
      dt$method <- tool
      return(dt)
    }))
    return(dt)
  }))
  return(predict_stats[!is.na(train_auc)])
}

plot_comp1 <- function(dt) {
  ggboxplot(dt, 
          x = 'method', y = 'test_auc', color = 'method', 
          add = 'jitter', legend = 'none') + 
    stat_compare_means(method = 'wilcox.test', paired = T, method.args = list('alternative' = 'greater'), 
                     comparisons = list(c('maui_factors', 'mofa_factors'), 
                                        c('maui_factors', 'pca_factors'))) +
    labs(y = 'AUC') + theme(axis.text = element_text(size = 10))
}

plot_comp2 <- function(dt) {
  dtc <- melt(dcast(dt, label ~ method, value.var = 'test_auc'), 
              id.vars = c('label', 'maui_factors'))
  dtc$difference <- dtc$maui_factors - dtc$value
  ggplot(dtc, 
         aes(x = maui_factors, y = value)) + 
    geom_point() + geom_abline(slope = 1) + 
    facet_wrap(~ variable, nrow = 1) + 
    labs(y = 'AUC')
} 

get_top_by_label <- function(dt, label.type, topN = 5) {
  do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
}

# find percent contribution of different datatypes to the top factor specific variables 
# LFs: Latent factors; nested list
# FWs: Feature weights for each LF; nested list 
# tool: pca/maui/mofa
# project: e.g. pancancer
# datatype: e.g. cnv_gex_mut
# factors: factor vector (e.g. cancer types for each sample or subtype info for each sample)
# top_lfs: top LF to pick per factor
# top_featuers: top features to pick per latent factor
plot_datatype_contribution_by_factor <- function(LFs, FWs, tool, project, datatype, 
                                                 factors, top_lfs = 1, top_features = 100) {
  experiment <- paste0(project, '.', datatype)
  L <- LFs[[tool]][[experiment]]
  W <- FWs[[tool]][[experiment]]
  dt <- get_factor_specific_variables(L, factors)
  top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', topN = top_lfs)
  # for each factor, pick top lfs and pick top features per lf and plot composition
  freq <- do.call(rbind, lapply(split(top_vars, top_vars$ref_cl), function(x) {
    l <- unique(x$variable) # one or more lfs 
    # get top features per lf and concatenate
    f <- do.call(c, lapply(l, function(y) {
      head(sort(abs(W[,y]), decreasing = T), top_features)
    }))
    freq <- data.table(table(sub("^(.+?)\\..+$", "\\1", unique(names(f)))))
    freq$factor <- unique(x$ref_cl)
    colnames(freq)[1] <- 'omics'
    return(freq)
  }))
  
  ggplot(freq, aes(x = factor, y = N)) + 
    geom_bar(stat = 'identity', aes(fill = omics), 
             position = 'fill') + coord_flip() + 
    labs(x = '', y = 'Percentage of top features contributing to top latent factors')
}
```

```{r pancancer_prediction, fig.height=9, fig.width=12}
# Find out if cancer types are predictable using latent factors 
pancancer_stats <- predict_labels(LFs, surv, tcga_subtypes, 'pancancer', 
                                  'cnv_gex_meth_mut', label.type = 'cancer_type', num.trees = 1000)
p1 <- plot_comp1(pancancer_stats) + theme(axis.text.x = element_blank())
p2 <- plot_comp2(pancancer_stats)
# plot cancer specific factors
M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
p3 <- plot_label_specific_vars(M, ct)

# data type contribution to top LF per factor
p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
                                           'pancancer', 'cnv_gex_meth_mut', factors = ct, 
                                           top_lfs = 1, top_features = 100) 
p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
  theme(legend.position = 'right', legend.background = element_blank())

Fig1 <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B')), 
                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('C', 'D')), 
                  nrow = 2,
                  rel_heights = c(2,3))

print(Fig1)
ggsave('pancancer_cancer_type_prediction.pdf', plot = Fig1, units = 'in', width = 12, height = 9)

# predictability of subtypes using pancancer LFs
pancancer_subtype_stats <- predict_labels(LFs, surv, tcga_subtypes, 'pancancer', 
                                  'cnv_gex_meth_mut', label.type = 'subtype', num.trees = 1000)

p1 <- plot_comp1(pancancer_subtype_stats) 
p2 <- plot_comp2(pancancer_subtype_stats)
# plot subtype specific factors
M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# remove NA values and pick subtypes with at least 20 samples 
selected <- names(which(table(st) > 20))
M <- M[st %in% selected,]
labels <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected

p3 <- plot_label_specific_vars(M, labels)

Fig2 <- plot_grid(plot_grid(p1, p2, nrow = 2), p3, ncol = 2, rel_widths = c(2, 3), 
          labels = 'AUTO')
print(Fig2)

ggsave('pancancer_subtype_prediction.pdf', plot = Fig2, units = 'in', width = 12, height = 9)

# train/test accuracy correlation 
#ggscatter(pancancer_subtype_stats, x = 'train_auc', y = 'test_auc', 
 #         color = 'method', facet.by = 'method', add = 'reg.line')
```

# Unsupervised pan-cancer clustering 

We do unsupervised clustering of samples and measure AMI to observe which tool's clusters correspond best to 
ground truth (cancer types)

```{r unsupervised_clustering, fig.height=12, fig.width=10}
get_plot <- function(M, plot.type = 'cancertype') {
  
  df <- plot_tsne(M)
  if(plot.type == 'cancertype') {
    df$tcga <- surv[match(rownames(df), bcr_patient_barcode)]$project
    df <- df[!is.na(df$tcga),]
  } else if (plot.type == 'subtype') {
    df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
    df <- df[!is.na(df$tcga),]
  }
  clust <- get_kmeans_subtypes(M[rownames(df),], k_vals = length(unique(df$tcga)), 0)
  df$cluster <- as.factor(clust[rownames(df),])


  p <- ggplot(df, aes(x = tSNE1, y = tSNE2))  
  p1 <- p + geom_point(aes(color = cluster)) +
    geom_text_repel(data = get_centroid(df[,1:2], df$cluster), 
              aes(x,y,label = text.label), size = 6) + 
    theme(legend.position = 'none')
  p2 <- p + geom_point(aes(color = tcga)) + 
          geom_text_repel(data = get_centroid(df[,1:2], df$tcga), 
              aes(x,y,label = gsub('TCGA-', '', text.label)), size = 3) +
    theme(legend.position = 'none')

  p <- cowplot::plot_grid(p1, p2)
  # also compute mutual information
  ami <- aricode::AMI(df$tcga, df$cluster)
  return(list('p' = p, 'ami' = ami))
}

pr <-  'pancancer' #'pancancer 99%' # 'nsclc 98%' pan-gastrointestinal 98% 'pan-gyn 98%'
e <- paste0(pr, '.cnv_gex_meth_mut')

#M <- LFs$pca_factors[[e]]
#ct <- surv[match(rownames(M), bcr_patient_barcode)]$project

# for maui, we drop redundant LFs 
p1 <- get_plot(remove_redundant_variables(LFs$maui_factors[[e]], perc = 99), 'cancertype') 
p2 <- get_plot(LFs$pca_factors[[e]], 'cancertype') # use all pcs, 
p3 <- get_plot(LFs$mofa_factors[[e]], 'cancertype') # mofa

# use AMI values to make another plot
p4 <- ggplot(melt(data.frame('MAUI' = p1$ami, 'PCA' = p2$ami, 'MOFA' = p3$ami)), 
       aes(x = variable, y = value)) + 
  geom_bar(aes(fill = variable), stat = 'identity') + 
  geom_text(aes(y = value + 0.05, label = round(value, 2))) + 
  labs(y = 'AMI', x = '') + 
  theme(legend.title = element_blank()) + coord_equal()


Fig <- cowplot::plot_grid(p1$p, p2$p, p3$p, p4, nrow = 4, labels = 'AUTO')
print(Fig)
ggsave("pancancer.unsupervised_clustering.pdf", Fig, units = 'in', width = 10, height = 12)
```


# clustering metrics of various cancers 

Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
```{r clustering_metrics, fig.width=10, fig.height=8}
multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      Mr <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
      M <- t(subset_features_by_variance(t(Mr), topN = 10))
     } else {
      M <- M[,head(colnames(M), 10),drop = F]
    }
    val <- c('internal') 
    if(ncol(M) > 2) {val <- c('internal', 'stability')}
    res <- clValid::clValid(obj = M,
                          nClust = 3:8,
                          clMethods = 'kmeans',
                          validation = val,
                          maxitems = 10000,
                          nstart = 1000)
    memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']])))
    colnames(memberships) <- paste0('k', colnames(memberships))
    stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
    colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
    return(list('clustering_stats' = stats, 'cluster_memberships' = memberships))
  })
})
parallel::stopCluster(cl)

cluster.stats <- do.call(rbind, lapply(names(clusters), function(tool) {
  do.call(rbind, lapply(names(clusters[[tool]]), function(experiment) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      datatypes <- strsplit(experiment, "\\.")[[1]][2]
      stats <- clusters[[tool]][[experiment]][['clustering_stats']]
      stats$cancer_type <- cancer_type
      stats$datatypes <- datatypes
      stats$method <- tool
      return(stats)
  }))
}))

# plot silhouette/stability score comparison
dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb]
p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = list(c('maui_factors', 'pca_factors'), 
                                        c('maui_factors', 'mofa_factors'))) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')


dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb]
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = list(c('maui_factors', 'pca_factors'), 
                                        c('maui_factors', 'mofa_factors'))) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

# combine both scores and plot for different values of k
dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
            cancer_type + method + k ~ metric, value.var = 'score')
dt$kv <- paste0('kmeans(k=',dt$k,')')
p3 <- ggplot(dt,
       aes(x = Silhouette, y = APN)) + 
        geom_point(aes(color = method), size = 3) + 
        facet_wrap(~ kv, ncol = 2) +
        labs(title = 'Multi-omics clustering: Silhouette vs APN')

Fig4 <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
                  p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))
print(Fig4)
ggsave('cluster_stability_metrics.pdf', plot = Fig4, units = 'in', 
       width = 10, height = 8)

```

# survival-predictive factors

I look for survival-predictive maui LFs, show how much c-index improves on top of 
clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.

```{r survival_factors, fig.width=12, fig.height=8}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
                  'PFI', 'PFI.time')

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    require(data.table)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
    res <- get_prognostic_covariates(clin, M)
    if(!is.null(res)) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      res$project <- cancer_type
      v <- res[order(pval)][1]$variable
      dat <- merge(clin, M[,v,drop = F], by = 'row.names')
      rownames(dat) <- dat$Row.names
      dat$Row.names <- NULL
      # compute hazard ratios 
      cox1 <- coxph(Surv(PFI.time,PFI) ~ ., 
              data=dat, x = TRUE)
      hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
      hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
      colnames(hr) <- c('hr', 'hr_low', 'hr_high')
      rownames(hr) <- v
      p1 <- survminer::ggforest(cox1, data = dat)
      
      # split variable into high/low to compute survival adjusted curves
      cp <- find_survival_cutpoint(dat[,v,drop=F], surv)
      dat[,v] <- ifelse(dat[,v] > cp, paste(v, 'high'), paste(v, 'low'))
      cox2 <- coxph(Surv(PFI.time,PFI) ~ ., 
                    data=dat, x = TRUE)
      p2 <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')
      
      # compute Cindex with all survival-predictive LFs and without LFs 
      vars <- head(res[order(pval)]$variable, 10)
      rsf_stats <- fit_RSF_clin(clin, M[,vars], ntree = 1000)
      
      return(list('stats' = res, 'p1' = p1, 'p2' = p2, 'hr' = hr, 
                  'Cindex_base' = rsf_stats$C_wo, 'Cindex_with_LFs' = rsf_stats$C_with))
    }
    return(NULL)
  })
parallel::stopCluster(cl)

plots <- sapply(simplify = F, names(surv.stats), function(x) {   
  p <- surv.stats[[x]][['p2']] 
  if(!is.null(p)) { 
    p + theme(axis.text.x = element_text(size = 8), 
              axis.text.y = element_text(size = 8), 
              axis.title.x = element_blank(), 
              axis.title.y = element_blank(), 
              legend.title = element_blank(),
              legend.position = c(0.5, 0.9),
              legend.text = element_text(size = 8), 
              legend.background = element_blank(), 
              legend.direction = 'vertical') 
    }
})
plots <- plots[!sapply(plots, is.null)]
p1 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
                   labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
                   label_size = 8)
print(p1)
ggsave(filename = 'surv.adjustedcurves.prog_lfs.pdf', plot = p1, width = 12, height = 8, units = 'in')

# plot hazard ratios for top LF per cancer type 
hr <- do.call(rbind, lapply(names(surv.stats), function(x) {
  if(is.null(surv.stats[[x]])) {return(NULL)}
  message(x)
  df <- surv.stats[[x]][['hr']]
  df$project <- unlist(strsplit(x, '\\.'))[[1]]
  return(df)
}))
hr$latent_factor <- rownames(hr)
# remove hr with Inf values
hr <- hr[!is.infinite(hr$hr_high),]

p2 <- ggplot(hr, aes(x = reorder(paste0(project, ' (', latent_factor, ')'), hr), y = log(hr))) +
  geom_bar(stat = 'identity') + 
  geom_errorbar(aes(ymin = log(hr_low), ymax = log(hr_high)), width = 0.2) + 
  #geom_text(aes(label = latent_factor, y = ifelse(hr > 1, log(hr+0.005), log(hr-0.005))), 
   #         nudge_x = 0.25) + 
  labs(x = '', y = 'log(hazard-ratio) of top latent factor per cancer\n(adjusted for tumor stage + age + gender)')+
  coord_flip()


# make a plot to show improvement in C index with LFs on top of clinical factors
ci <- do.call(rbind, lapply(names(surv.stats), function(x) {
  if(is.null(surv.stats[[x]])) {return(NULL)}
  df <- data.frame('base' = surv.stats[[x]][['Cindex_base']], 
                   'with_lf' = surv.stats[[x]][['Cindex_with_LFs']], 
                   'project' = unlist(strsplit(x, '\\.'))[[1]])
  return(df)
}))

dt <- melt(ci, id.vars = 'project')
dt$variable <- ifelse(dt$variable == 'base', 'Only clinical factors', 'Clinical factors + MAUI LFs')

p3 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
  geom_segment(aes(x = reorder(project, with_lf), xend = project, 
                   y = base, yend = with_lf)) + 
  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 6) +
  coord_flip() + 
  labs(x = '', y = '') + 
  theme(legend.title = element_blank(), legend.position = 'bottom', 
        legend.background = element_blank(), 
        legend.margin = margin(), 
        legend.direction = 'vertical', legend.box.background = element_blank()) +
  scale_color_brewer(type = 'qual', palette = 6)

p <- cowplot::plot_grid(p3, p2, nrow = 1, labels = 'AUTO')

print(p)
ggsave(filename = 'surv.Cindex_improv.hazard_ratio.pdf', plot = p, width = 12, height = 8, units = 'in')

```

# clinical factor associated MAUI LFs

Find maui LFs associated to clinical factors and plot them

```{r clinical_factors, fig.width=12, fig.height=9}
experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
# find clinical-factor associated LFs 
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('experiments', 'LFs'))
clin.stats <- do.call(rbind, 
        pbapply::pbsapply(cl = cl, simplify = F, 
               experiments, function(experiment) {
               require(data.table)
               source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
               M <- LFs$maui_factors[[experiment]]
               clin <- process_clinical_covariates(rownames(M))
               dt <- do.call(rbind, 
                    pbapply::pbsapply(simplify = F, 
                                      colnames(clin), function(x) {
                      if(!is.numeric(clin[,x])) {
                        factors <- clin[,x]
                        m <- M[!is.na(factors),]
                        factors <- factors[!is.na(factors)]
                        dt <- get_factor_specific_variables(m, factors) #compute_anova(m, factors)
                        dt$label <- x
                        return(dt)
                      }
                      return(NULL)
                      }))
               dt$project <- unlist(strsplit(experiment, "\\."))[[1]]
               return(dt)
             }))
parallel::stopCluster(cl)

# find top factors per project per label and plot
dt <- clin.stats[padj < 0.05, .SD[which.min(padj)],by = c('project', 'label')]

require(ggrepel)
dt$label <- stringr::str_wrap(gsub('_', ' ', dt$label), width = 8)
p1 <- ggplot(dt, aes(x = project, y = label)) + 
  geom_point(aes(color = -log10(padj), size = -log10(padj))) + 
  geom_text_repel(aes(label = variable)) + 
  coord_flip() +
  scale_color_gradient(low = 'blue', high = 'red') + 
  labs(x = '', y = 'clinical factor')

# pick example cases and show more detailed LF activation 
# top examples by clinical factor type
# top 5 variables per clinical factor 

top <- clin.stats[padj < 0.05,.SD[which.min(padj)], by = label][order(padj),.SD[which.min(padj)],by = project]
plots <- sapply(simplify = FALSE, 1:4, function(i) {
  pr <- top$project[i]
  l <- top$label[i]
  M <- LFs$maui_factors[[paste0(pr, '.cnv_gex_meth_mut')]]
  factors <- process_clinical_covariates(rownames(M))[[l]]
  M <- M[!is.na(factors),]
  factors <- factors[!is.na(factors)]
  plot_label_specific_vars(M, factors, string_width = 24)+ #, top_vars = top_vars) + 
    labs(title = pr, subtitle = l) + 
    theme(legend.position = 'none')
})

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
p <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
print(p)
ggsave(filename = 'clinical_factors.LFs.pdf', p, units = 'in', width = 12, height = 9)
```

# Predicting MSI status using latent factors
```{r msi_stats, fig.width=10, fig.height=8}
# predict MSI status for pan-gastrointestinal tumors
# read MSI data
dataDir <- '/data/local/buyar/pancancer_diagnostics/data/GDCdata_prepared'
MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
  f <- file.path(dataDir, paste0(pr, '.MSI.tsv'))
  dt <- data.table::fread(f)[,c(2,4)]
  colnames(dt)[2] <- 'MSI'
  dt$project <- pr
  return(dt)
}))
# remove samples with undertermined MSI status
MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]
msi_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
  y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, partition = 0.6, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  #res <- predict_labels_glmnet(M, y)
  res$method <- tool
  return(res)
}))
msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))
# with glmnet and error bars 
# for supplements (includes other tools)
# p1 <- ggplot(msi_stats, aes(x = method, y = auc)) + 
#   geom_bar(aes(fill = method), stat = 'identity') + 
#   geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2) + 
#   labs(y = 'AUC',x = '') + theme(legend.title = element_blank()) + 
#   facet_grid(~ label)

p1 <- ggplot(msi_stats[method == 'MAUI'], aes(x = label, y = auc)) +
  geom_bar(stat = 'identity', aes(fill = label)) + 
  labs(y = 'AUC', x = '') +
  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2)+
  theme(legend.title = element_blank())

# make some maui specific plots to explore MSI related LFs
M <- remove_redundant_variables(LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']], 
                                perc = 99) # make it consistent with pan-cancer
y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI 
M <- M[!is.na(y),]
y <- y[!is.na(y)]

# tsne plot of top LF related to MSI-H and MSI classes
dt <- get_factor_specific_variables(M, y)
# use top predictive lfs to guide tsne 
top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)[['variable']])

df <- plot_tsne(M[,top_vars], y, returnData = T)
# color tsne by top lf informative for MSI-H
top_lf <- dt[order(padj)][ref_cl == 'MSI-H']$variable[1]
df <- merge(df, data.frame(M[,top_lf,drop =F]), by = 'row.names')
df$cancertype <- surv[match(df$Row.names, bcr_patient_barcode)]$project
df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected

p2.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
  theme(legend.title = element_blank())
p2.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes_string(color = top_lf)) + 
  scale_color_gradient(low = 'gray', high = 'red')
p2.3 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(cancertype))) + 
  theme(legend.title = element_blank())
p2.4 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(subtype))) + 
  theme(legend.title = element_blank())

print(cowplot::plot_grid(p2.1, p2.2, p2.3, p2.4, ncol = 2, labels = 'AUTO'))
# top features that contribute to top factors relevant to MSI status prediction
f <- data.table::fread(file.path(inputDir, 'maui_factors', 
                               paste('pan-gastrointestinal', # project 
                                      'cnv_gex_meth_mut', #data type
                                      'maui_factors', # tool
                                      'feature_weights.csv', sep = '.')))
feat.M <- as.matrix(f[,-1])
rownames(feat.M) <- gsub('assay: ', '', f$V1)

# sort features by contribution to LF and plot by data type 
df <- data.frame(feat.M[,top_lf,drop = F])
df$data.type <- sub('^(.+?)\\..+$', '\\1', rownames(df))

# get top 100 features
df <- df[order(abs(df[,top_lf]), decreasing = T),]
freq <- do.call(rbind, lapply(0:29, function(i) {
  s <- (i*100+1)
  e <- (i+1)*100
  x <- data.frame(table(df[s:e,]$data.type))
  x$rank <- stringr::str_wrap(paste0("[",s,",",e,"]"),
                              width = 8)
  x$ind <- i
  return(x)
}))
p3 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
  geom_bar(aes(fill = Var1), stat='identity') + theme(legend.title = element_blank()) + 
  labs(x = paste0("top features contributing to ",top_lf)) 
  

p <- cowplot::plot_grid(p1, p3, p2.1, p2.2, labels = 'AUTO')
print(p)
ggsave(filename = 'MSI.status.pdf', p, units = 'in', width = 10, height = 8)

```

# Semi-supervised clustering of samples 

1. Cluster samples by LFs associated to known TCGA subtypes or those that are associated to survival outcomes. 
2. Annotate each cluster using cancer state scores or hallmark gene set scores 

## TCGA subtype informed clusters 

### Use-case: pan-gastrointestinal cancers
```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
require(survminer)
require(cluster)
tcga_subtypes <- process_tcga_subtypes()
# subtype specific LFs
get_top_by_label <- function(dt, label.type, topN = 5) {
  do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
}

pr <- 'pan-gastrointestinal' #'TCGA-BRCA' # 'glioma' 
experiment <- paste0(pr, '.cnv_gex_meth_mut')
M <- LFs$maui_factors[[experiment]]
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# find subtype specific LFs 
dt <- get_factor_specific_variables(M, st)

# use gap statistic to get clusters
vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 3)[['variable']])
# use gap statistic to find optimum clusters 
#r1 <- cluster::clusGap(M[,vars], FUNcluster = kmeans, K.max = 20, iter.max = 50, B = 100)
#plot(r1)
#k <- cluster::maxSE(r1$Tab[,3], r1$Tab[,4], method = 'firstSEmax', SE.factor = 1)
k <- length(na.omit(unique(st)))
# cluster samples using only subtype specific factors 
clust <- data.frame(get_kmeans_subtypes(M[,vars], k_vals = k, min_cluster_size = 0))

# plots
df <- plot_tsne(M[,vars], as.factor(clust[,1]), returnData = T)
colnames(df)[3] <- 'maui'
df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
df$cancertype <- surv[match(rownames(df), bcr_patient_barcode)]$project

p <- ggplot(df, aes(x = tSNE1, y = tSNE2))  
p1 <- p + geom_point(aes(color = maui)) +
  geom_text_repel(data = get_centroid(df[,1:2], df$maui), 
            aes(x,y,label = text.label), size = 6)
p2 <- p + geom_point(aes(color = tcga)) + 
        geom_text_repel(data = get_centroid(df[,1:2], df$tcga), 
            aes(x,y,label = text.label), size = 5)
p3 <- plot_survival(surv, df[,'maui',drop =F])[['plot']] + theme(legend.position = 'none') + labs(y = 'Progression Free Survival')
p4 <- plot_survival(surv, df[,'tcga',drop =F])[['plot']] + theme(legend.position = 'none') + labs(y = 'Progression Free Survival')

# use hallmark gene sets to annotate cluster specific functions
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))
S <- scores[intersect(rownames(M), rownames(scores)),]
clust$tcga <- tcga_subtypes[match(rownames(clust), bcr_patient_barcode)]$Subtype_Selected

# plot top gene sets
h1 <- plot_label_specific_vars(S, clust[rownames(S), 1], 
                              as_heatmap = T, cellwidth = 10, 
                              cellheight = 12,
                              top_vars_per_label = 2)
h2 <- plot_label_specific_vars(S, clust[rownames(S), 2], 
                              as_heatmap = T, cellwidth = 10, 
                              cellheight = 12, 
                              top_vars_per_label = 2)

# altogether
P <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
                         labels = 'AUTO', ncol = 2)
print(P)
ggsave(filename = paste0('semi_supervised.clustering.', experiment,'.pdf'), 
       P, units = 'in', 
       width = 10, height = 12)

```

## Clustering informed by survival outcomes

Pick survival-predictive LFs and cluster samples based on those
```{r semi_supervised_survival, fig.width=10, fig.height=12}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
                  'PFI', 'PFI.time')

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
                                function(experiment){
    require(data.table)
    require(survminer)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
    res <- get_prognostic_covariates(clin, M)
    # now cluster samples based on only survival predictive factors
    prog_lfs <- unique(res$variable)
    if(length(prog_lfs) < 2) {
      return(NULL)
    }
    clust <- get_spectral_clusters(M[,prog_lfs], k_vals = 2:10)
    # find value of k which minimizes log-rank p-value
    dt <- do.call(rbind, lapply(colnames(clust), function(x) {
      message(x)
      pval <- plot_survival(surv, clust[,x,drop=F], return_pval = T)
      data.table('k' = x, 'pval' = pval)
    }))
    
    # return clustering with best result
    top <- dt[order(pval)][1]$k
    return(list('prog.lf' = M[,prog_lfs], 'prog.clust' = clust[,top,drop = F]))
})
parallel::stopCluster(cl)

# compare maui prognostic clusters with TCGA labels 
stats <- do.call(rbind, lapply(names(prog.clust), function(experiment) {
  message(experiment)
  if(is.null(prog.clust[[experiment]])) {return(NULL)}
  df <- prog.clust[[experiment]][['prog.clust']]
  colnames(df)[1] <- 'maui'
  df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
  do.call(rbind, lapply(colnames(df), function(x) {
    if(length(unique(df[,x])) < 2) {
      pval <- NA
    } else {
      pval <- plot_survival(surv, df[,x,drop = F], return_pval = T)
    }
    return(data.table('source' = x, 'pval' = pval, 'experiment' = experiment))
  }))
}))

stats$cancertype <- gsub('.cnv_gex_meth_mut', '', stats$experiment)
# which cancers can we improve upon tcga?
p1 <- ggplot(stats, aes(x = cancertype, y = -log10(pval))) +  
  geom_bar(stat = 'identity', aes(fill = source), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  coord_flip() + 
  scale_fill_brewer(type = 'qual', palette = 6) 

# use hallmark gene sets to annotate cluster specific functions
#scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))

# use canonical pathways to annotate prognostic groups
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_canonical_pathways/ssgsea.canonical_pathways.RDS'))


# scores geneset scores
# M survival predictive LF matrix
# maui_clust: best survival prognostic clustering of M 
plot_clusts <- function(M, maui_clust, scores) {
  df <- plot_tsne(M, as.factor(maui_clust[,1]), returnData = T)
  colnames(df)[3] <- 'maui'
  #df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
  
  p1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
    geom_point(aes(color = maui)) +
    geom_text_repel(data = get_centroid(df[,1:2], df$maui), 
            aes(x,y,label = text.label), size = 6)
  # plot survival for the worst prognostic group, and the rest
  ms <- get_median_survival_by_group(surv, df[,'maui',drop=F])

  # find clusters with worst/best prognosis (by median survival)
  worst <- names(sort(ms)[1])
  best <- names(sort(ms, T)[1])
  
  p2 <- plot_survival(surv, 
                      df[df$maui %in% c(best, worst), 
                         'maui', drop = F])[['plot']] +
    labs(y = 'PFI') +
    theme(legend.position = 'left', legend.title = element_blank()) + 
    scale_color_brewer(type = 'qual', palette = 6)
  
  # heatmap of hallmark scores for best/worst survivors
  selected <- intersect(rownames(df[df$maui %in% c(worst, best),]), rownames(S))
  dt <- get_factor_specific_variables(S[selected,], df[selected, 'maui'])
  # select factors 
  dt[padj < 0.05][order(padj)]
  h1 <- plot_label_specific_vars(S[selected,], 
                                 df[selected, 'maui'], 
                                 padj_threshold = 0.1,
                                as_heatmap = T, cellwidth = 15, 
                                top_vars_per_label = 5, cellheight = 15)
  return(list('p1' = p1, 'p2' = p2, 'h1' = h1))
}

# plot examples for which we improve upon tcga
dt <- dcast(stats, experiment ~ source, value.var = 'pval')
dt$improvement <- -log10(dt$maui) - -log10(dt$tcga)
dt <- dt[order(improvement, decreasing = T)]

pr <- 'TCGA-UCEC'
e <- paste0(pr, '.cnv_gex_meth_mut')
M <- prog.clust[[e]]$prog.lf
maui_clust <- prog.clust[[e]]$prog.clust
S <- scores[intersect(rownames(M), rownames(scores)),]
P <- plot_clusts(M, maui_clust, S)

Pc <- cowplot::plot_grid(p1, plot_grid(P$p1, P$p2, nrow = 1, labels = c('B', 'C')), 
                   P$h1$gtable, nrow = 3, labels = c('A', '', 'D'))

print(Pc)
ggsave(filename = paste0('semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
       Pc, units = 'in', 
       width = 10, height = 12)

```


# Classify metastatic samples

Use TCGA-SKCM as case study to see if we can predict which samples are metastatic and which ones are primary
We have multi-omics data from 360 metastatic tumors and 100 primary tumors (distinct group of patients)
```{r, eval = FALSE}
# read batch info to see which patient samples are metastatic
batches <- do.call(rbind, lapply(dir(assay_dir, "TCGA-SKCM.*info.tsv", full.names = T),
                  data.table::fread))
# get sample types from sample barcodes
st <- get_tcga_sample_types(batches$V1)
st$bcr_patient_barcode <- sub("^(.{12}).+$", "\\1", st$barcode)
# primary tumors: TP, metastatic tumors: TM
st <- unique(st[sampletype %in% c('TP', 'TM')][,2:3])

skcm_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['TCGA-SKCM.cnv_gex_meth_mut']]
  y <- st[match(rownames(M), bcr_patient_barcode)]$sampletype
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  # plot_tsne(M, y)
  res$method <- tool
  return(res)
}))

# with glmnet 
ggplot(skcm_stats[label == 'TM'], aes(x = method, y = auc)) + 
  geom_bar(aes(fill = method), stat = 'identity') + 
  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2) + 
  labs(y = 'AUC',x = '') + theme(legend.title = element_blank()) 

# plot maui LFs predictive of metastatic/primary tumor
M <- LFs$maui_factors[['TCGA-SKCM.cnv_gex_meth_mut']]
y <- st[match(rownames(M), bcr_patient_barcode)]$sampletype
M <- M[!is.na(y),]
y <- y[!is.na(y)]
dt <- get_factor_specific_variables(M, y)
top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', topN = 5)[['variable']]

plot_tsne(M[,top_vars], y)

df <- data.frame(M[,top_vars])
df$y <- as.factor(y)
fit <- ranger::ranger(seed = 42, formula = y ~ ., data = df, probability = T, num.trees = 1000)

cp <- data.frame(fit$predictions)
rownames(cp) <- rownames(df)
cp$type <- st[match(rownames(cp), bcr_patient_barcode)]$sampletype

# find metastatic samples that look like primary tumor samples
d1 <- cp[cp$type == 'TM' & cp$TM > 0.7,]
d1$confident <- 'confident_metastatic'
d2 <- cp[cp$type == 'TM' & cp$TM < 0.7,]
d2$confident <- 'not_confident_metastatic'
plot_survival(surv, rbind(d1, d2)[,'confident',drop = F], duration_col = 'OS.time',
              event_col = 'OS')


# check cancer state scores to see if unconfident metastatic samples have similar scores as primary ones
dataDir <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore/msigdb_canonical_pathways/'
files <- dir(dataDir, '.scores.csv', full.names = T)
scores <- do.call(rbind, sapply(simplify = F, files, function(x) {
 data.table::fread(x)
}))
colnames(scores)[1] <- 'bcr_patient_barcode'

cp$group <- 'primary'
cp[cp$type == 'TM' & cp$TM >= 0.7,]$group <- 'confident_metastatic'
cp[cp$type == 'TM' & cp$TM < 0.7,]$group <- 'not_confident_metastatic'

# get cancer state scores
dt <- scores[match(rownames(cp), bcr_patient_barcode),]
M <- as.matrix(dt[,-1])
rownames(M) <- dt$bcr_patient_barcode

dt <- get_factor_specific_variables(M, as.factor(cp$group))

B <- get_basis_matrix(M, cp$group)
#B2 <- apply(B, 2, function(x) x/max(x)+max(x)/4)
#pheatmap(B2)
pheatmap(B[,unique(dt[padj < 0.05]$variable)])


# what are the genes that distinguish confident metastatic from 
M <- LFs$maui_factors$`TCGA-SKCM.cnv_gex_meth_mut`
plot_label_specific_vars(M, cp$group, top_vars_per_label = 5)
dt <- get_factor_specific_variables(M, cp$group)
top <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)

```

# Show subtypes depend on the objective

Patients clustering depend on the objective of stratification

```{r}
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')

project <- 'pan-gastrointestinal'
datatype <- 'cnv_gex_meth_mut'
experiment <- paste0(project, '.', datatype)

M <- LFs$maui_factors[[experiment]]
M <- t(subset_features_by_variance(t(remove_redundant_variables(M)), topN = 20))

clin <- process_clinical_covariates(rownames(M))
clin <- subset(clin, select = intersect(colnames(clin), clin_factors))

pfi_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'PFI', duration_col = 'PFI.time')
pfi_clust <- get_kmeans_subtypes(M[,unique(pfi_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
pfi_k <- names(which.min(sapply(colnames(pfi_clust), function(x) plot_survival(surv, pfi_clust[,x,drop = F], 'PFI.time', 'PFI', T))))

os_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'OS', duration_col = 'OS.time')
os_clust <- get_kmeans_subtypes(M[,unique(os_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
os_k <- names(which.min(sapply(colnames(os_clust), function(x) plot_survival(surv, os_clust[,x,drop = F], 
                                                                             'OS.time', 'OS', T))))

# combine with known pan-gi subtype annotation

df <- merge(pfi_clust[,pfi_k,drop = F], os_clust[,os_k,drop = F], by = 'row.names')
df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected

tsne <- plot_tsne(remove_redundant_variables(M), returnData = T)

df$tSNE1 <- tsne[df$Row.names,]$tSNE1
df$tSNE2 <- tsne[df$Row.names,]$tSNE2

mdf <- melt(df, id.vars = c('Row.names', 'tSNE1', 'tSNE2'))

ggplot(mdf, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes(color = as.factor(value))) + 
  facet_wrap(~ variable, nrow = 1)
```



# predict mutational signatures
```{r}
# import mutational signature data
mutsig <- data.table::fread('/data/local/buyar/pancancer_diagnostics/data/mutational_signatures/signature_profile_sample.txt')
mutsig$Tumor_Sample_Barcode <- gsub("\\.", "-", mutsig$Tumor_Sample_Barcode)
# only keep TCGA samples
mutsig <- mutsig[grep('TCGA', Tumor_Sample_Barcode)]
dt <- data.table::dcast(mutsig, Tumor_Sample_Barcode ~ Signature, value.var = 'Contribution')

cl <- parallel::makeCluster(30) 
parallel::clusterExport(cl, varlist = c('dt', 'LFs'))
res <- do.call(rbind, pbapply::pbsapply(simplify = F, cl = cl, unique(mutsig$Signature), function(x) {
  require(data.table)
  source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
  r <- do.call(rbind, lapply(names(LFs), function(m) {
    message(m)
    M <- LFs[[m]]$pancancer.cnv_gex_meth_mut
    s <- predict_numerical_glmnet(M, dt[match(rownames(M), Tumor_Sample_Barcode)][[x]])
    s$tool <- m
    return(s)
  }))
  r$sig <- x
  return(r)
}))
parallel::stopCluster(cl)

ggplot(res, aes(x = reorder(sig, Rsquare), y = Rsquare)) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') + 
  coord_flip() + labs(x = '')

df <- melt(dcast(res, sig ~ tool, value.var = 'Rsquare'), 
                id.vars = c('sig', 'maui_factors'))

ggplot(df, aes(x = maui_factors, y = value)) + geom_point() + 
  facet_wrap(~ variable, nrow = 1) + 
  geom_abline(slope = 1)
```

# predictability of clinical factors 

```{r, eval = FALSE}
# predict tumor purity using latent factors 
res <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pancancer.cnv_gex_meth_mut']]
  y <- process_clinical_covariates(rownames(M))[,'tumor_purity_IHC']
  res <- predict_numerical_glmnet(M, y)
  res$method <- tool
  return(res)
}))

# predict tumor stage for pan-gi 
pangi_stats <- predict_labels(LFs, surv, tcga_subtypes, 
                              'pan-gastrointestinal', 'cnv_gex_meth_mut', label.type = 'clinical')
ggplot(pangi_stats[label.type == 'ajcc_pathologic_tumor_stage'][label != 'Undefined'], aes(x = label, y = test_auc)) + 
  geom_bar(aes(fill = method), stat = 'identity', position = 'dodge') +
  coord_flip()

# tumor stage 
label <- 'age_at_initial_pathologic_diagnosis'
experiment <- 'TCGA-BRCA.cnv_gex_meth_mut'
stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][[experiment]]
  y <- process_clinical_covariates(rownames(M))[[label]]
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y)
  res <- predict_numerical_glmnet(M, y)
  res$method <- tool
  return(res)
}))
# with glmnet multi-label pred
ggplot(stats, aes(x = method, y = Accuracy)) + 
  geom_bar(stat = 'identity', aes(fill = method)) + 
  geom_errorbar(aes(ymin = AccuracyLower, ymax = AccuracyUpper), width = 0.2)

```


# predictability of cancer states

```{r, eval = F}
# read cancer sea gene set scores
#dataDir <- '/data/local/buyar/pancancer_diagnostics/data/geneset_scores/cancersea'
#files <- dir(dataDir, '.scores.csv', full.names = T)
#scores <- do.call(rbind, sapply(simplify = F, files[1:3], function(x) {
#  data.table::fread(x)
#}))
#colnames(scores)[1] <- 'bcr_patient_barcode'

scores <- readRDS('/data/local/buyar/arcas/subtyping_paper/data/mogsa_scores/MOGSA_scores.RDS')
scores <- do.call(rbind, lapply(scores$cnv_gex_meth_mut, function(x) data.table(x, keep.rownames = T)))
colnames(scores)[1] <- 'bcr_patient_barcode'

# elastic net regression on pathway scores usign LFs as input
predict_scores <- function(scores, LFs, label, experiment)
  do.call(rbind, lapply(names(LFs), function(tool) {
    message(tool)
    M <- LFs[[tool]][[experiment]]
    y <- scores[match(rownames(M), bcr_patient_barcode)][[label]]
    M <- M[!is.na(y),]
    y <- y[!is.na(y)]
    res <- predict_numerical_glmnet(M, y)
    res$method <- tool
  return(res)
}))

labels <- c('ANGIOGENESIS', 'GLYCOLYSIS', 'COAGULATION', 'DNA_REPAIR', 
            'HYPOXIA', 'ESTROGEN_RESPONSE_EARLY', 'ESTROGEN_RESPONSE_LATE' ,
            'EPITHELIAL_MESENCHYMAL_TRANSITION')
brca_scores <- do.call(rbind, lapply(labels, function(l) {
  message(l)
  df <- predict_scores(scores, LFs, l, 'TCGA-BRCA.cnv_gex_meth_mut')
  df$label <- l
  return(df)
}))

ggplot(brca_scores, aes(x = label, y = Rsquare)) + 
  geom_bar(stat = 'identity', aes(fill = method), position = 'dodge') + 
  coord_flip()
```




# pan-cancer survival factors 
```{r, eval = FALSE}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
                  'PFI', 'PFI.time')
experiment <- 'pancancer.cnv_gex_meth_mut'
res <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][[experiment]]
  clin <- process_clinical_covariates(rownames(M))
  clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
  cl <- parallel::makeCluster(50)
  parallel::clusterExport(cl, varlist = c('clin', 'M'))
  res <- do.call(rbind, pbapply::pblapply(cl = cl, 1:100, function(i) {
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    require(data.table)
    s <- sample(rownames(M), 0.5 * nrow(M))
    dt <- get_prognostic_covariates(clin[s,], M[s,])
    dt$i <- i
    return(dt)
  }))
  parallel::stopCluster(cl)
  res$method <- tool
  return(res)
}))

# get top factor per method
plots <- sapply(simplify = F, names(LFs), function(tool) {
  M <- LFs[[tool]][[experiment]]
  top_fac <- res[method == tool][,length(unique(i)),by = 'variable'][order(V1, decreasing = T)]$variable[1:3]
  cowplot::plot_grid(plotlist = lapply(top_fac, function(x) {
     p <- plot_survival(surv, data.frame('cl' = ifelse(M[,x] > median(M[,x]), 
                                             'high', 'low')))
     p$plot
  }), nrow = 1)
})
dev.off()
cowplot::plot_grid(plotlist = plots, nrow = 3, labels = names(plots))

M <- LFs[[tool]][[experiment]]
plot_tsne(M, M[,'LF32'])
```



<!-- Cluster samples into k = number of cancer types in the project -->

<!-- ```{r} -->
<!-- cl <- parallel::makeCluster(20) -->
<!-- parallel::clusterExport(cl, varlist = c('LFs', 'projectDict')) -->
<!-- clusters <- sapply(simplify = F, names(LFs), function(tool) { -->
<!--   message("Clustering top ",tool) -->
<!--   pbapply::pbsapply(cl = cl, simplify = F, names(LFs[[tool]]), function(experiment) { -->
<!--     message(experiment) -->
<!--     project <- strsplit(experiment, '\\.')[[1]][1] -->
<!--     datatype <- strsplit(experiment, '\\.')[[1]][2] -->
<!--     source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R') -->
<!--     require(data.table) -->
<!--     M <- LFs[[tool]][[experiment]] -->
<!--     if(tool == 'maui_factors') { -->
<!--       M <- t(subset_features_by_variance(t(M), topN = 10)) -->
<!--     } else { -->
<!--       M <- M[,head(colnames(M), 10),drop = F] -->
<!--     } -->
<!--     val <- c('internal')  -->
<!--     #if(ncol(M) > 2) {val <- c('internal', 'stability')} -->
<!--     #k <- length(projectDict[[project]]) -->
<!--     res <- clValid::clValid(obj = M, -->
<!--                           nClust = 3:10, -->
<!--                           clMethods = 'kmeans', -->
<!--                           validation = val, -->
<!--                           maxitems = 10000, -->
<!--                           nstart = 1000) -->
<!--     memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']]))) -->
<!--     colnames(memberships) <- paste0('k', colnames(memberships)) -->
<!--     stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')] -->
<!--     colnames(stats) <- c('metric', 'k', 'clustering_method', 'score') -->
<!--     return(list('clustering_stats' = stats, 'cluster_memberships' = memberships)) -->
<!--   }) -->
<!-- }) -->
<!-- parallel::stopCluster(cl) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- stats <- do.call(rbind, lapply(names(clusters), function(tool) { -->
<!--   do.call(rbind, lapply(names(clusters[[tool]]), function(experiment) { -->
<!--       cancer_type <- strsplit(experiment, "\\.")[[1]][1] -->
<!--       datatypes <- strsplit(experiment, "\\.")[[1]][2] -->
<!--       stats <- clusters[[tool]][[experiment]][['clustering_stats']] -->
<!--       stats$cancer_type <- cancer_type -->
<!--       stats$datatypes <- datatypes -->
<!--       stats$method <- tool -->
<!--       return(stats) -->
<!--   })) -->
<!-- })) -->
<!-- ``` -->

<!-- clustering metrics -->
<!-- ```{r} -->
<!-- ggboxplot(stats[datatypes == 'cnv_gex_mut'], x = 'method', y = 'score',  -->
<!--           add = 'jitter', facet.by = 'metric', color = 'method') -->
<!-- ``` -->

<!-- Find out how much the cluster labels correspond to cancer types  -->
<!-- ```{r} -->
<!-- cl <- parallel::makeCluster(20) -->
<!-- parallel::clusterExport(cl, varlist = c('clusters', 'projectDict', 'surv', )) -->
<!-- # compute mutual information stats (between cluster labels and cancer types) -->
<!-- MI_stats <- do.call(rbind, sapply(simplify = F, names(clusters), function(tool) { -->
<!--   message("computing MI for ",tool) -->
<!--   do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, names(clusters[[tool]]), -->
<!--                                    function(experiment) { -->
<!--     message(experiment) -->
<!--     project <- strsplit(experiment, '\\.')[[1]][1] -->
<!--     datatype <- strsplit(experiment, '\\.')[[1]][2] -->
<!--     source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R') -->
<!--     require(data.table) -->
<!--     k <- stats[method == tool][cancer_type == project][datatypes == datatype][,.SD[which.max(score)]]$k -->
<!--     clust <- clusters[[tool]][[experiment]][['cluster_memberships']][,paste0('k',k),drop=F] -->
<!--     labels <- surv[match(rownames(clust), bcr_patient_barcode)]$project -->

<!--     data.table('method' = tool, 'project' = project, 'datatypes' = datatype,  -->
<!--                'cramersV' = compute_cramers_V(cbind(clust, labels))[['cramersV']],  -->
<!--                'ARI' = aricode::ARI(clust[,1], labels),  -->
<!--                'AMI' = aricode::AMI(clust[,1], labels)) -->
<!--                                    })) -->
<!--   })) -->
<!-- parallel::stopCluster(cl) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ggboxplot(MI_stats[datatypes == 'cnv_gex_mut'],  -->
<!--           x = 'method', y = 'AMI',  -->
<!--           add = 'jitter') +  -->
<!--     stat_compare_means(method = 'wilcox.test', paired = T, method.args = list('alternative' = 'greater'),  -->
<!--                      comparisons = list(c('maui_factors', 'mofa_factors'),  -->
<!--                                         c('maui_factors', 'pca_factors'))) + -->
<!--   labs(title = 'One-vs-all test-AUC by cancer type') -->

<!-- ``` -->


<!-- Unsupervised clustering experiment -->
<!-- ```{r} -->
<!-- get_plot <- function(M, plot.type = 'cancertype') { -->

<!--   df <- plot_tsne(M) -->
<!--   if(plot.type == 'cancertype') { -->
<!--     df$tcga <- surv[match(rownames(df), bcr_patient_barcode)]$project -->
<!--   } else if (plot.type == 'subtype') { -->
<!--     df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected -->
<!--   } -->
<!--   clust <- get_kmeans_subtypes(M[rownames(df),], k_vals = length(unique(df$tcga)), 0) -->
<!--   df$maui <- as.factor(clust[rownames(df),]) -->


<!--   p <- ggplot(df, aes(x = tSNE1, y = tSNE2))   -->
<!--   p1 <- p + geom_point(aes(color = maui)) + -->
<!--     geom_text_repel(data = get_centroid(df[,1:2], df$maui),  -->
<!--               aes(x,y,label = text.label), size = 6) -->
<!--   p2 <- p + geom_point(aes(color = tcga)) +  -->
<!--           geom_text_repel(data = get_centroid(df[,1:2], df$tcga),  -->
<!--               aes(x,y,label = text.label), size = 5) -->

<!--   p <- cowplot::plot_grid(p1, p2) -->
<!--   # also compute mutual information -->
<!--   ami <- aricode::AMI(df$tcga, df$maui) -->
<!--   return(list('p' = p, 'ami' = ami)) -->
<!-- } -->

<!-- pr <- 'pancancer' #'pancancer' # 'pan-gastrointestinal' -->
<!-- e <- paste0(pr, '.cnv_gex_meth_mut') -->
<!-- M <- LFs$maui_factors[[e]] -->
<!-- ct <- surv[match(rownames(M), bcr_patient_barcode)]$project -->
<!-- st <-  tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected -->

<!-- dt.ct <- get_factor_specific_variables(M, ct) -->
<!-- ct_vars <- unique(get_top_by_label(dt.ct[padj < 0.05][order(padj)], label.type = 'ref_cl', 2)[['variable']]) -->
<!-- dt.st <- get_factor_specific_variables(M, st) -->
<!-- st_vars <- unique(get_top_by_label(dt.st[padj < 0.05][order(padj)], 'ref_cl', 2)[['variable']]) -->

<!-- # LFs with most variance -->
<!-- top_vars <- rownames(subset_features_by_variance(t(M), topN = 42)) -->
<!-- # using top vars will cause the clusters dominated by the major component to dominate -->

<!-- p1 <- get_plot(M, 'cancertype') # use all factors  (unsupervised) -->
<!-- p2 <- get_plot(M[,top_vars], 'cancertype') # use top most variable factors (unsupervised) -->
<!-- p3 <- get_plot(M[,ct_vars], 'cancertype') # use cancer type specific factor (supervised ) -->
<!-- p4 <- get_plot(remove_redundant_variables(M, cutoff = 0.6), 'cancertype') # (unsupervised) -->
<!-- p5 <- get_plot(LFs$pca_factors[[e]]) # pca -->
<!-- p5.1 <- get_plot(LFs$pca_factors[[e]][,1:10]) # pca -->
<!-- p5.2 <- get_plot(LFs$pca_factors[[e]][,1:20]) # pca -->
<!-- p5.4 <- get_plot(LFs$pca_factors[[e]][,1:40]) -->

<!-- p6 <- get_plot(LFs$mofa_factors[[e]]) # mofa -->

<!-- ps <- lapply(c(5:8/10), function(r) { -->
<!--   get_plot(remove_redundant_variables(M, cutoff = r), 'cancertype') -->
<!-- }) -->

<!-- dt <- do.call(rbind, lapply(experiments, function(e) { -->
<!--   x <- LFs$maui_factors[[e]] -->
<!--   y <- remove_redundant_variables(x, cutoff = 0.6) -->
<!--   data.table('before' = ncol(x), 'after' = ncol(y), 'n' = nrow(x), 'exp' = e,  -->
<!--              'redundant' = ncol(x) - ncol(y)) -->
<!-- })) -->

<!-- p5 <- ggscatter(dt, x = 'n', y = 'redundant', add = 'reg.line') + -->
<!--    stat_cor(label.x = 400, label.y = 80, size = 6) + -->
<!--   labs(x = 'Number of samples', y = 'Percentage of redundant LFs') + -->
<!--   stat_regline_equation(label.x = 400, label.y = 75, size = 6) +  -->
<!--   geom_text_repel(aes(label = gsub('.cnv_gex_meth_mut', '', exp))) -->

<!-- ``` -->


